\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}

\title{SIMC-Shape Pion-Background Subtraction Procedure}
\author{KaonLT Production Analysis}
\date{\today}

\begin{document}
\maketitle

\section{Purpose}

This note documents the current \texttt{simc\_shape\_components} pion-background
subtraction procedure implemented in the production analysis.
The goal is to replace the legacy single-scale pion subtraction with an
MM-dependent weighting constructed from SIMC component fits.

The three pion-background components are
\begin{itemize}[leftmargin=2em]
  \item $\pi$-$n$,
  \item $\pi$-$\Delta$,
  \item $\pi$-SIDIS.
\end{itemize}

The procedure is implemented through the following code paths:
\begin{itemize}[leftmargin=2em]
  \item \texttt{src/cuts/pion\_component\_fits.py},
  \item \texttt{src/cuts/pion\_component\_subtraction.py},
  \item \texttt{src/cuts/rand\_sub.py},
  \item \texttt{src/binning/calculate\_yield.py},
  \item \texttt{src/binning/ave\_per\_bin.py},
  \item \texttt{src/utility/background\_config.py}.
\end{itemize}

\section{Definitions}

Let the unit-normalized SIMC MM templates be
\begin{align}
  T_n(MM), \qquad
  T_\Delta(MM), \qquad
  T_{\mathrm{SIDIS}}(MM).
\end{align}

The pion-control fitted model is
\begin{align}
  P_{\mathrm{ctrl}}^{\mathrm{fit}}(MM)
  =
  B_n T_n(MM)
  +
  B_\Delta T_\Delta(MM)
  +
  B_{\mathrm{SIDIS}} T_{\mathrm{SIDIS}}(MM),
\end{align}
and the kaon-side fitted pion-background model is
\begin{align}
  K_{\pi\mathrm{-bg}}^{\mathrm{fit}}(MM)
  =
  A_n T_n(MM)
  +
  A_\Delta T_\Delta(MM)
  +
  A_{\mathrm{SIDIS}} T_{\mathrm{SIDIS}}(MM).
\end{align}

The event-by-event pion weight is defined as
\begin{align}
  w_\pi(MM)
  =
  \frac{K_{\pi\mathrm{-bg}}^{\mathrm{fit}}(MM)}
       {P_{\mathrm{ctrl}}^{\mathrm{fit}}(MM)}.
\end{align}

The weighted pion-control model is therefore
\begin{align}
  w_\pi(MM)\,P_{\mathrm{ctrl}}^{\mathrm{fit}}(MM)
  =
  K_{\pi\mathrm{-bg}}^{\mathrm{fit}}(MM),
\end{align}
which is the model-level closure identity.

In the current tuning scheme, the pion-control post-fit component scales are
kept fixed at unity so that \(P_{\mathrm{ctrl}}^{\mathrm{fit}}(MM)\) remains
the unmodified control-model denominator. Any phenomenological post-fit scaling
is applied only on the kaon-side pion-background model
\(K_{\pi\mathrm{-bg}}^{\mathrm{fit}}(MM)\).

\section{Part 1: Load SIMC Component Shapes}

The SIMC component shapes are loaded from existing ROOT files and converted into
unit-normalized MM templates.

\subsection{Inputs}

The current mode is controlled in
\texttt{src/utility/background\_config.py} through
\begin{itemize}[leftmargin=2em]
  \item \texttt{PARTICLE\_SUBTRACTION\_MODE\_DEFAULT},
  \item \texttt{SIMC\_TREE\_NAME\_DEFAULT},
  \item \texttt{SIMC\_PION\_COMPONENT\_BACKGROUND\_MAP}.
\end{itemize}

The default SIMC tree name is
\begin{align}
  \texttt{h10}.
\end{align}

\subsection{Shape construction}

For each component:
\begin{enumerate}[leftmargin=2em]
  \item open the corresponding SIMC ROOT file,
  \item read the SIMC tree,
  \item loop over events,
  \item fill an MM histogram using the SIMC event weight,
  \item apply the same SIMC-side cuts used for MM-shape comparisons,
  \item normalize the resulting MM histogram to unit area.
\end{enumerate}

The loader can produce:
\begin{itemize}[leftmargin=2em]
  \item a setting-wide full-range MM template,
  \item optional binned templates in $(t,\phi)$.
\end{itemize}

If a fine $(t,\phi)$ component template is empty or has non-usable support, the
analysis can fall back hierarchically to:
\begin{enumerate}[leftmargin=2em]
  \item the direct $(t,\phi)$ template,
  \item the corresponding $t$-aggregated template,
  \item the setting-wide template.
\end{enumerate}

\subsection{MM alignment}

When the target data side uses an MM offset, the SIMC component templates are
shifted by the same data-side MM offset before fitting. This is handled inside
\texttt{build\_particle\_subtraction\_component\_result(...)} so that the fit is
performed with aligned MM locations.

\section{Part 2: Fit the SIMC Shapes}

\subsection{Two fit targets}

The templates are fit to two different histograms:
\begin{itemize}[leftmargin=2em]
  \item the pion-control data histogram,
  \item the kaon no-subtraction histogram.
\end{itemize}

For binned yield and averaging paths, the fit first attempts the direct
$(t,\phi)$ component template for each channel. If that template is empty or
non-usable, the code falls back to a broader support level as described in
Part~1 rather than treating ordinary sparse-bin behavior as a fatal physics
failure.

The pion-control fit determines
\begin{align}
  B_n,\quad B_\Delta,\quad B_{\mathrm{SIDIS}},
\end{align}
while the kaon-side fit determines
\begin{align}
  A_n,\quad A_\Delta,\quad A_{\mathrm{SIDIS}}.
\end{align}

\subsection{Ordered staged fit}

The staged fit order is controlled directly in
\texttt{PARTICLE\_SUBTRACTION\_COMPONENT\_FIT\_WINDOW\_CONFIG}.

Current defaults are:
\begin{itemize}[leftmargin=2em]
  \item pion-control:
  \begin{align}
    \pi\mathrm{-}n
    \;\rightarrow\;
    \pi\mathrm{-SIDIS}
    \;\rightarrow\;
    \pi\mathrm{-}\Delta,
  \end{align}
  \item kaon no-sub:
  \begin{align}
    \pi\mathrm{-}n
    \;\rightarrow\;
    \pi\mathrm{-}\Delta
    \;\rightarrow\;
    K\mathrm{-}\Sigma^0
    \;\rightarrow\;
    \pi\mathrm{-SIDIS}.
  \end{align}
\end{itemize}

Each stage fits the residual remaining after the previous fitted components have
been subtracted from the input histogram. Conceptually:
\begin{align}
  R_0(MM) &= D(MM), \\
  R_1(MM) &= D(MM) - C_1 T_1(MM), \\
  R_2(MM) &= D(MM) - C_1 T_1(MM) - C_2 T_2(MM),
\end{align}
where $D(MM)$ is the fit target and the $C_i$ are the staged component
amplitudes in the configured fit order.

The anchor windows for the current defaults are stored in
\texttt{PARTICLE\_SUBTRACTION\_COMPONENT\_FIT\_WINDOW\_CONFIG}.

\subsection{How the amplitudes are found}

At each staged step, the code solves for a single non-negative amplitude against
the current residual histogram. The exact rule depends on the configured
\texttt{stage\_amplitude\_mode}.

\subsubsection*{Window-integral mode}

If a component is configured with
\texttt{stage\_amplitude\_mode = window\_integral}, then the amplitude is taken
from the ratio of the residual integral to the unit-normalized template
integral inside the configured stage-amplitude window:
\begin{align}
  C
  =
  \max\!\left(
    0,\,
    \frac{\sum\limits_{i \in W_{\mathrm{amp}}} R(i)}
         {\sum\limits_{i \in W_{\mathrm{amp}}} T(i)}
  \right),
\end{align}
where
\begin{itemize}[leftmargin=2em]
  \item $R(i)$ is the residual bin content,
  \item $T(i)$ is the template bin content,
  \item $W_{\mathrm{amp}}$ is the configured stage-amplitude window.
\end{itemize}

This is a direct window-normalization estimate. The implementation then stores
the resulting amplitude uncertainty as the residual quadrature error inside the
same window divided by the template integral in that window.

\subsubsection*{Least-squares mode}

If a component is configured with
\texttt{stage\_amplitude\_mode = least\_squares}, the code minimizes the
single-template weighted residual over the selected fit bins:
\begin{align}
  \chi^2(C)
  =
  \sum_i
  \frac{\left[R(i) - C\,T(i)\right]^2}{\sigma_i^2}.
\end{align}

For one template, the solution is
\begin{align}
  C
  =
  \max\!\left(
    0,\,
    \frac{\sum_i T(i)\,R(i)/\sigma_i^2}
         {\sum_i T(i)^2/\sigma_i^2}
  \right).
\end{align}

The reported one-parameter uncertainty is
\begin{align}
  \delta C
  =
  \left(
    \sum_i \frac{T(i)^2}{\sigma_i^2}
  \right)^{-1/2}.
\end{align}

So in practical terms:
\begin{itemize}[leftmargin=2em]
  \item \texttt{window\_integral} uses an area match inside a core window,
  \item \texttt{least\_squares} uses the actual template shape against the
  residual.
\end{itemize}

\subsubsection*{Residual update}

After a stage amplitude $C_j$ is found, the cumulative staged model is updated
and the next component is fit to the new residual:
\begin{align}
  R_j(MM)
  =
  D(MM)
  -
  \sum_{k=1}^{j} C_k T_k(MM).
\end{align}

The amplitudes are explicitly constrained to remain non-negative at every
staged step.

\subsubsection*{Joint refinement after the staged pass}

Once the ordered staged amplitudes are available, the code can refine all
component amplitudes simultaneously with a non-negative bounded linear least
squares solve. The staged solution provides the starting point and prior target.

The refined objective is equivalent to minimizing
\begin{align}
  \chi^2_{\mathrm{data}}
  +
  \sum_j
  \left(
    \frac{C_j - C^{\mathrm{stage}}_j}{\tau_j}
  \right)^2,
\end{align}
where $\tau_j$ is the configured prior width for component $j$.

The implementation solves this as an augmented linear system with
\begin{align}
  C_j \ge 0
\end{align}
for all components.

\subsubsection*{Coordinate-update fallback}

If the joint prior refinement is not accepted by the validation logic, the code
can perform ordered coordinate updates. For a given component $j$, it forms the
residual with all other components held fixed:
\begin{align}
  R_{-j}(i)
  =
  D(i) - \sum_{k \ne j} C_k T_k(i),
\end{align}
and updates
\begin{align}
  C_j
  =
  \max\!\left(
    0,\,
    \frac{\sum_i T_j(i)\,R_{-j}(i)/\sigma_i^2}
         {\sum_i T_j(i)^2/\sigma_i^2}
  \right),
\end{align}
with the prior penalty included in the numerator and denominator when the prior
width is finite.

\subsection{Current default windows}

\subsubsection*{Pion-control fit}
\begin{align}
  \pi\mathrm{-}n &: [0.90,\,0.94], \\
  \pi\mathrm{-}\Delta &: [1.16,\,1.25], \\
  \pi\mathrm{-SIDIS} &: [1.36,\,1.45].
\end{align}

There is also a current stage-amplitude core window for $\pi$-$\Delta$:
\begin{align}
  [1.17,\,1.23].
\end{align}

\subsubsection*{Kaon no-sub fit}
\begin{align}
  \pi\mathrm{-}n &: [0.90,\,0.94], \\
  \pi\mathrm{-}\Delta &: [1.18,\,1.26], \\
  \pi\mathrm{-SIDIS} &: [1.36,\,1.45].
\end{align}

The kaon-side fit currently excludes the configured Sigma region
\begin{align}
  [1.18,\,1.23]
\end{align}
from the relevant fit objects used for validation and staged diagnostics.

\subsection{Post-fit component scales}

After the staged amplitudes are extracted, the code can apply component-specific
post-fit scale factors from the config.

Current defaults:
\begin{center}
\begin{tabular}{@{}lccc@{}}
\toprule
Fit target & $\pi$-$n$ & $\pi$-$\Delta$ & $\pi$-SIDIS \\
\midrule
Pion control & 0.95 & 0.95 & 0.75 \\
Kaon no-sub  & 0.95 & 0.95 & 0.25 \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Refinement and validation}

The current implementation also supports:
\begin{itemize}[leftmargin=2em]
  \item joint refinement with Gaussian-like prior penalties around the staged solution,
  \item ordered coordinate updates,
  \item full-range validation checks,
  \item over-subtraction diagnostics.
\end{itemize}

The fallback mode is currently set to
\begin{align}
  \texttt{error},
\end{align}
so component mode is expected to fail hard rather than silently revert to the
legacy subtraction path.

\section{Part 3: Build and Apply the Event Weight}

\subsection{Histogram-level weight construction}

Once the fitted amplitudes are available, the code constructs:
\begin{itemize}[leftmargin=2em]
  \item \texttt{H\_pion\_control\_model},
  \item \texttt{H\_kaon\_pion\_model},
  \item \texttt{H\_pion\_weight\_vs\_MM},
  \item \texttt{H\_weighted\_pion\_control\_model}.
\end{itemize}

The MM weight is evaluated bin by bin as
\begin{align}
  w_i = \frac{K_{\pi\mathrm{-bg},i}^{\mathrm{fit}}}
               {P_{\mathrm{ctrl},i}^{\mathrm{fit}}},
\end{align}
subject to the configured denominator floor and clip bounds.

\subsection{Event-template construction}

The actual subtraction template is not the weighted model.
It is built by looping over the pion-control event source:
\begin{itemize}[leftmargin=2em]
  \item prompt,
  \item random,
  \item dummy,
  \item dummy-random,
\end{itemize}
with the usual signs and normalization coefficients, and applying
\begin{align}
  w_\pi(MM_{\mathrm{event}})
\end{align}
to each accepted event.

This produces the weighted pion-control event template that is actually
subtracted from the kaon data histograms.

\subsection{Subtraction}

For the relevant analysis scope, the corrected kaon histogram is formed as
\begin{align}
  H_{\mathrm{kaon,after}}
  =
  H_{\mathrm{kaon,before}}
  -
  H_{\mathrm{weighted\ pion\ template}}.
\end{align}

This same logic is applied in:
\begin{itemize}[leftmargin=2em]
  \item \texttt{rand\_sub.py},
  \item \texttt{calculate\_yield.py},
  \item \texttt{ave\_per\_bin.py},
\end{itemize}
with scope-specific histograms and caches.

\section{Diagnostic Interpretation}

\subsection{Model closure}

The page
\begin{quote}
\texttt{Part 3 model closure: weighted pion model vs kaon bg model}
\end{quote}
checks the algebraic identity
\begin{align}
  w_\pi(MM)\,P_{\mathrm{ctrl}}^{\mathrm{fit}}(MM)
  =
  K_{\pi\mathrm{-bg}}^{\mathrm{fit}}(MM).
\end{align}

If this page fails, the problem is in the weight construction itself:
binning mismatch, normalization inconsistency, or an implementation bug.

\subsection{Event-template closure}

The page
\begin{quote}
\texttt{Part 3 event-template closure vs kaon bg model}
\end{quote}
checks whether the weighted \emph{actual pion-control events} reproduce the
kaon-side fitted pion-background model.

This is the physically relevant closure test because the weighted event template
is what is actually subtracted from kaon data.

If model closure is good but event-template closure is poor, then the weight
formula is behaving correctly while the fitted pion-control model does not map
cleanly onto the real pion-control event distribution.

\subsection{Kaon data vs pion-background models}

The page
\begin{quote}
\texttt{Part 3 kaon data vs pion-background models}
\end{quote}
is a full-range comparison of:
\begin{itemize}[leftmargin=2em]
  \item the kaon data before pion subtraction,
  \item the fitted kaon-side pion-background model,
  \item the weighted pion-control event template.
\end{itemize}

This page is intended only to show how large the subtraction source is relative
to the full kaon spectrum. It is not itself the closure test.

\subsection{Before/after subtraction}

The page
\begin{quote}
\texttt{Part 3 full-range MM before/after pion subtraction}
\end{quote}
shows the full-range kaon histogram before subtraction, the weighted pion
template, and the resulting after-subtraction histogram.

\section{Current Main Configuration Knobs}

The principal controls for the current implementation are:
\begin{itemize}[leftmargin=2em]
  \item \texttt{PARTICLE\_SUBTRACTION\_MODE\_DEFAULT},
  \item \texttt{PARTICLE\_SUBTRACTION\_FALLBACK\_MODE\_DEFAULT},
  \item \texttt{PARTICLE\_SUBTRACTION\_WEIGHT\_DENOM\_FLOOR\_DEFAULT},
  \item \texttt{PARTICLE\_SUBTRACTION\_WEIGHT\_CLIP\_MIN\_DEFAULT},
  \item \texttt{PARTICLE\_SUBTRACTION\_WEIGHT\_CLIP\_MAX\_DEFAULT},
  \item \texttt{PARTICLE\_SUBTRACTION\_WEIGHT\_WARN\_MAX\_DEFAULT},
  \item \texttt{PARTICLE\_SUBTRACTION\_WINDOW\_CONFIG},
  \item \texttt{PARTICLE\_SUBTRACTION\_COMPONENT\_FIT\_WINDOW\_CONFIG}.
\end{itemize}

\section{Summary}

The procedure has three logically separate steps:
\begin{enumerate}[leftmargin=2em]
  \item load aligned unit-area SIMC component shapes;
  \item fit those shapes to pion-control and kaon no-sub histograms;
  \item form an MM-dependent event weight and apply it to the actual pion-control events.
\end{enumerate}

The key distinction is:
\begin{itemize}[leftmargin=2em]
  \item \emph{weighted pion-control model}: an algebra/implementation check,
  \item \emph{weighted pion event template}: the actual subtraction object.
\end{itemize}

For physics diagnosis, the most important page is the event-template closure page,
because it tests whether the weighted pion-control event sample really reproduces
the kaon-side fitted pion background that the subtraction is meant to remove.

\end{document}
